---
title: "CUMTD Data Collection"
author: "Shaw Kagawa"
output: 
  html_document:
    theme: sandstone
    highlight: tango
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DBI)
library(magrittr)
library(tidyverse)
library(plotly)
library(reshape)
library(hms)
db_file <- "stop_times.db"
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)

stop_times <- tbl(con, "stop_times", stringAsFactors=FALSE)

colors <- read_csv("google_transit/routes.txt") %>%
        select(route_id, route_color)

routeToCol <- Vectorize(function(route_id) {
        paste0("#", colors$route_color[colors$route_id==route_id])
})

convertToSeconds <- function (X) {
        X <- strsplit(X, ":")
        sapply(X, function(Y) sum(as.numeric(Y) * c(3600, 60, 1)))
}
```

## Data Exploration {.tabset}
### Format
<h3>Format</h3>
(Collected from CUMTD's API- see [CUMTD Developer Resources](https://developer.cumtd.com))
```{r, echo=FALSE}
renderTable({
        stop_times %>% 
                select(trip_id:route_id, `2018-12-03`:`2018-12-06`) %>% 
                as.tibble() %>%
                sample_n(5)
})
```
The data has observations for each departure in the CUMTD system, with a trip_id (unique
to each sequence of stops within the route), an arrival_time, a stop_id 
(which shows which stop the arrival is at), and a route_id (the route displayed on the 
headsign of the bus: 'ILLINI', 'GOLDHOPPER', etc.).  
The consecutive columns are appended as each day's trips are logged, and is an integer 
showing how many seconds the specific arrival was late or early (negative integer meaning
the arrival was early, positive meaning the arrival was late).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
all_times <- stop_times %>%
        select(route_id, arrival_time) %>%
        collect() %>%
        mutate(start = arrival_time,
               end = start) %>%
        select(route_id, start, end) %>%
        arrange(start) %>%
        group_by(route_id) %>%
        summarise(min_start=min(start), max_end=max(end)) %>%
        mutate(route_color=routeToCol(route_id),
               route=str_extract(route_id, "(?!([0-9]| )).[A-Z]+")) %>%
        group_by(route) %>%
        arrange(desc(min_start), .by_group=TRUE)

ggplotly(all_times %>% {
        ggplot(.) + 
                geom_rect(aes(xmin=convertToSeconds(.$min_start), 
                              xmax=convertToSeconds(.$max_end), 
                              ymin=row(.)[,1]/2-0.2, 
                              ymax=row(.)[,1]/2+0.2,
                              text=paste("Route:", .$route_id, 
                                         "<br>Start Time:", .$min_start,
                                         "<br>End Time:", .$max_end)),
                          fill=.$route_color,
                          color="black",
                          size=0.05) +
                labs(x='Time of Day', title='Running routes by time of day') +
                theme(axis.title.y=element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank()) +
                scale_x_time(breaks=as.hms(c(paste0(seq(6,33,3), ":00:00"))))
})
```
The buses run from 6am (06:00) to 6am (30:00), and the graph shows this relationship. 
Each color represents each different bus route, ranging from BLUE to YELLOWHOPPER.
  

### Delays by Bus Route
We can graph all of the delays by bus route and sort them by average delay on a specific 
day.
```{r, echo=FALSE}
dateInput("dateInput2", label="Choose a Date:", 
          min="2018-12-01",
          max=tail(colnames(stop_times), n=1), 
          value=tail(colnames(stop_times), n=1))

renderPlotly({
        date <- as.character(as.Date(input$dateInput2))
        ggplotly(stop_times %>%
                group_by(route_id) %>%
                select(date) %>%
                summarise_all(mean, na.rm=TRUE) %>%
                collect() %>%
                filter(.[[date]] != "NA") %>%
                mutate(route_color=routeToCol(route_id),
                       minutes=.[[date]] / 60) %>%
                arrange(minutes) %>% {
                ggplot(.) +
                        geom_col(aes(x=fct_reorder(.$route_id, .$minutes), y=.$minutes,
                                     text=paste("Bus Route:", .$route_id,
                                              "<br>Seconds:", round(.$minutes * 60, 1),
                                              "<br>Minutes:", round(.$minutes, 2))),
                                 fill=.$route_color) +
                        coord_flip() +
                        labs(x="Bus Route", 
                             y="Average Delay (minutes) (<0 = early, >0 = late)",
                             title=paste("Average Bus Delay by Bus Route on", date))
                }, tooltip="text")
})
```

But maybe this may be inaccurate due to error bars.
 

### Delays by Date
We can graph the delays by bus route by day, and find any days where delays might be 
an outlier.
```{r, echo=FALSE}
renderUI({
        selectInput(inputId="busRoute", 
                    label="Choose a Bus Route:", 
                    choices=getBusRoutes(),
                    selected="ILLINI")
})
   
renderPlotly({
        stop_times %>%
                filter(route_id==input$busRoute) %>%
                select(`2018-12-01`:as.character(Sys.Date())) %>%
                summarise_all(mean, na.rm=TRUE) %>%
                as.tibble() %>%
                gather(date, delay) %>%
                mutate(date=lubridate::ymd(date), delay=delay/60) %>%
                ggplot(aes(x=date, y=delay)) +
                        geom_col() +
                        scale_x_date() +
                        labs(x="Date", 
                             y="Average Delay (minutes)",
                             title=paste("Average Delay by Day for", input$busRoute))
})
```

##
***

## Hypothesis Testing {.tabset}
### Significant Bus Routes
We can figure out if a certain bus route has a significantly different mean delay 
compared to the other bus routes.

Here are summary stats for all bus routes ordered by number of departures:
```{r, echo=FALSE}
dateInput("dateInput3",
          label="Choose a Date:",
          min="2018-12-01", max=tail(colnames(stop_times), n=1),
          value=tail(colnames(stop_times), n=1))

DT::renderDataTable({ 
        date <- as.character(as.Date(input$dateInput3))
        dateColumn <- as.symbol(date)
        stop_times %>%
                group_by(route_id) %>%
                select(route_id, date) %>%
                mutate(delay=dateColumn) %>%
                filter(delay != "NA") %>%
                summarise(num_departures=n(),
                          mean=mean(delay, na.rm=TRUE), 
                          sample_sd=sd(delay, na.rm=TRUE)) %>%
                arrange(desc(num_departures), mean) %>%
                as.data.frame()
})
```

Kruskal-Wallis test on the bus route delays:
```{r, echo=FALSE}
renderPrint({
        date <- as.character(as.Date(input$dateInput3))
        dateColumn <- as.symbol(date)
        data <- stop_times %>%
                group_by(route_id) %>%
                select(date) %>%
                mutate(delay=dateColumn) %>%
                filter(delay != "NA") %>%
                select(delay) %>%
                as.data.frame()
        data$route_id <- as.factor(data$route_id)
        with(data, kruskal.test(delay, route_id))
})
```

Pairwise Wilcox Mann-Whitney U Test on bus delays to find significantly different routes. 
This shows the number of other bus routes that were significantly different in their delay 
distribution compared to each other bus route.
```{r, echo=FALSE}
renderPlotly({
        date <- as.character(as.Date(input$dateInput3))
        dateColumn <- as.symbol(date)
        data <- stop_times %>%
                group_by(route_id) %>%
                select(date) %>%
                mutate(delay=dateColumn) %>%
                filter(delay != "NA") %>%
                select(delay) %>%
                as.data.frame()
        data$route_id <- as.factor(data$route_id)
        test <- with(data, pairwise.wilcox.test(delay, route_id, 
                                        p.adjust="bonferroni"))
        ggplotly(melt(test[[3]]) %>%
                filter(value != "NA") %>%
                group_by(X2) %>%
                summarise(n=sum(value < 0.05)) %>%
                mutate(route_color=routeToCol(X2)) %>%
                arrange(n) %>% {
                ggplot(.) +
                        geom_col(aes(x=fct_reorder(.$X2, .$n), y=.$n,
                                 text=paste("Route:", .$X2,
                                            "<br>n:", .$n)),
                             fill=.$route_color) +
                        coord_flip() +
                        labs(x="Route", 
                             y="Number of significantly different routes to this route",
                             title="Measure of how different each route is compared to other routes") +
                        coord_flip() +
                        geom_hline(aes(text="\"Significance Level\" (half the # of routes)",
                                       yintercept=length(levels(data$route_id)) / 2,
                                       col="RED")) +
                        theme(legend.position="none")
                }, tooltip="text")
})
```
The next question is figuring out what level would be considered significantly different 
from all the other hours. This "significance level" was set to half the size of the sample 
for now.


### Significant Hour on Given Date
```{r, echo=FALSE}
getBusRoutes <- reactive({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        routes <- stop_times %>%
                        group_by(route_id) %>%
                        select(route_id, date) %>%
                        mutate(delay=dateColumn) %>%
                        filter(delay != "NA") %>%
                        ungroup() %>%
                        distinct(route_id) %>%
                        pull()
        return(routes)
})

inputPanel(
        dateInput("dateInput4",
                  label="Choose a Date:",
                  min="2018-12-01", max=tail(colnames(stop_times), n=1),
                  value=tail(colnames(stop_times), n=1)),
        renderUI({
                selectInput(inputId="busRoute2", 
                            label="Choose a Bus Route:", 
                            choices=getBusRoutes(),
                            selected="ILLINI")
        })
)
```

We can figure out if some delays are unexpectedly high at certain times of day.
``` {r, echo=FALSE}
renderPlotly({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                select(arrival_time, date) %>%
                collect() %>%
                mutate(hour=stringr::str_sub(arrival_time, 1, 2)) %>%
                group_by(hour) %>%
                summarise(n=n()) %>%
                ggplot(aes(x=hour, y=n)) + 
                        geom_col() +
                        labs(x="Hour of Day", y="Number of Departures",
                             title="Number of Departures by Hour")
})

renderPlotly({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        hours <- stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                pull(arrival_time) %>% 
                stringr::str_sub(1, 2) %>% 
                as.factor()
        delays <- stop_times %>%
                filter(route_id==input$busRoute2) %>%
                select(date) %>%
                mutate(delay=dateColumn / 60) %>%
                filter(delay != "NA") %>%
                select(delay)
        data.frame(delay=delays, hour=hours) %>%
                ggplot(aes(x=hour, y=delay)) + 
                        geom_violin() +
                        labs(x="Hour of Day", y="Delay (minutes)",
                             title="Delays by Hour")
})
```

Is there a statistical difference between the distributions of these times? Since these don't 
seem to be normally distributed, we can use a nonparametric test to figure this out. (If this 
fails, then there are probably too little data points to compare)
```{r, echo=FALSE}
renderPrint({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        hours <- stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                pull(arrival_time) %>% 
                stringr::str_sub(1, 2) %>% 
                as.factor()
        delays <- stop_times %>%
                filter(route_id==input$busRoute2) %>%
                select(date) %>%
                mutate(delay=dateColumn / 60) %>%
                filter(delay != "NA") %>%
                pull(delay)
        data <- data.frame(delay=delays, hour=hours)
        with(data, kruskal.test(delay, hour))
})
```

A Pairwise Wilcox Mann-Whitney U Test on bus delays can find significantly different times 
of day where buses have a different distribution of delays. The table below shows the 
results of the test, but because of the varying sample sizes (see graph above), we cannot 
really compare these hours as equal.
```{r, echo=FALSE}
renderPlotly({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        hours <- stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                pull(arrival_time) %>% 
                stringr::str_sub(1, 2) %>% 
                as.factor()
        delays <- stop_times %>%
                filter(route_id==input$busRoute2) %>%
                select(date) %>%
                mutate(delay=dateColumn / 60) %>%
                filter(delay != "NA") %>%
                pull(delay)
        data <- data.frame(delay=delays, hour=hours)
        test <- with(data, pairwise.wilcox.test(delay, hour, p.adjust.method = "bonferroni"))
        ggplotly(melt(test[[3]]) %>%
                group_by(X2) %>%
                filter(value != "NA") %>%
                summarise(n=sum(value < 0.05)) %>%
                mutate(hour=X2) %>%
                ggplot(aes(x=hour, y=n, text=paste("Hour:", hour,
                                                   "<br>n:", n))) +
                        geom_col() +
                        labs(x="Hour of Day", 
                             y="Number of significantly different hours to this hour",
                             title="Measure of how different each hour is compared to other hours", 
                             subtitle="Pairwise Wilcox Test") +
                        geom_hline(aes(text="\"Significance Level\" (half the # of observations)",
                                       yintercept=length(levels(hours)) / 2, col="RED")) +
                        theme(legend.position="none")
        , tooltip="text")
})
```
The next question is figuring out what level would be considered significantly different 
from all the other hours. This "significance level" was set to half the size of the sample, 
for now.

