---
title: "CUMTD Data Collection"
author: "Shaw Kagawa"
output: 
  html_document:
    theme: united
    highlight: tango
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DBI)
library(magrittr)
library(tidyverse)
library(plotly)
library(reshape)
db_file <- "stop_times.db"
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)

stop_times <- tbl(con, "stop_times")
```

## Data Exploration {.tabset}
### Format
<h3>Format</h3>
(Collected from CUMTD's API- see [CUMTD Developer Resources](https://developer.cumtd.com))
```{r, echo=FALSE}
renderTable({
        stop_times %>% 
                select(trip_id:route_id, `2018-12-03`:`2018-12-06`) %>% 
                as.tibble() %>%
                sample_n(5)
})
```
The data has observations for each departure in the CUMTD system, with a trip_id (unique
to each sequence of stops within the route), an arrival_time, a stop_id 
(which shows which stop the arrival is at), and a route_id (the route displayed on the 
headsign of the bus: 'ILLINI', 'GOLDHOPPER', etc.).  
The consecutive rows are appended as each day's trips are logged, and is an integer 
showing how many seconds the specific arrival was late or early (negative integer meaning
the arrival was early, positive meaning the arrival was late).

```{r, echo=FALSE}
all_times <- stop_times %>%
        select(route_id, arrival_time) %>%
        collect() %>%
        mutate(group = str_extract(route_id, "(?!([0-9]| )).[A-Z]+"),
           start = arrival_time,
           end = start) %>%
        select(group, route_id, start, end) %>%
        arrange(start) %>%
        group_by(route_id) %>%
        slice(c(1, n())) %>%
        as.data.frame()
for (i in 1:nrow(all_times)) {
        if (mod(i, 2) == 1) {
                all_times[i, 4] <- all_times[i+1, 4]
        }
}

convertToSeconds <- function(X) {
        X <- strsplit(X, ":")
        sapply(X, function(Y) sum(as.numeric(Y) * c(3600, 60, 1)))
}

arranged_times <- all_times %>% 
        mutate(index=row_number()) %>% 
        filter(mod(index, 2) == 1) %>% 
        mutate(start=convertToSeconds(start),
               end=convertToSeconds(end))

plot <- arranged_times %>% 
        ggplot() + 
        geom_rect(aes(xmin=start, xmax=end, 
                      ymin=index/2-0.45, ymax=index/2+0.45, 
                      color=route_id), fill=NA) +
        labs(x='Time of Day', y='Index', title='Running routes by time of day') +
        scale_x_time()

ggplotly(plot)
```
The buses run from 6am (06:00) to 6am (30:00), and the graph shows this relationship. 
Each color represents each different bus route, ranging from BLUE to YELLOWHOPPER.
  

### Delays by Bus Route
We can graph all of the delays by bus route and sort them by average delay on a specific 
day.
```{r, echo=FALSE}
dateInput("dateInput2", label="Choose a Date:", min="2018-12-01", max=Sys.Date(), value=Sys.Date())

renderPlotly({ 
        date <- as.character(as.Date(input$dateInput2))
        dateColumn <- as.symbol(date)
        stop_times %>%
                group_by(route_id) %>%
                select(date) %>%
                summarise_all(mean, na.rm=TRUE) %>%
                filter(dateColumn != "NA") %>%
                mutate(minutes=dateColumn / 60) %>%
                ggplot(aes(x=fct_reorder(route_id, desc(minutes)), y=minutes)) +
                        geom_col() +
                        coord_flip() +
                        labs(x="Bus Route", 
                             y="Average Delay (minutes) (<0 = early, >0 = late)",
                             title=paste("Average Bus Delay by Bus Route on", date))
})
```

But maybe this may be inaccurate due to error bars.
 

### Delays by Date
We can graph the delays by bus route by day, and find any days where delays might be 
an outlier.
```{r, echo=FALSE}
renderUI({
        selectInput(inputId="busRoute", 
                    label="Choose a Bus Route:", 
                    choices=getBusRoutes(),
                    selected="ILLINI")
})
   
renderPlotly({
        stop_times %>%
                filter(route_id==input$busRoute) %>%
                select(`2018-12-01`:as.character(Sys.Date())) %>%
                summarise_all(mean, na.rm=TRUE) %>%
                as.tibble() %>%
                gather(date, delay) %>%
                mutate(date=lubridate::ymd(date), delay=delay/60) %>%
                ggplot(aes(x=date, y=delay)) +
                        geom_col() +
                        scale_x_date() +
                        labs(x="Date", 
                             y="Average Delay (minutes)",
                             title=paste("Average Delay by Day for", input$busRoute))
})
```

##
***

## Hypothesis Testing {.tabset}
### Delays by Bus Route
We can figure out if a certain bus route has a significantly different mean delay 
compared to the other bus routes.

Here are summary stats for all bus routes ordered by number of departures:
```{r, echo=FALSE}
dateInput("dateInput3",
          label="Choose a Date:",
          min="2018-12-01", max=Sys.Date(),
          value=Sys.Date())

DT::renderDataTable({ 
        date <- as.character(as.Date(input$dateInput3))
        dateColumn <- as.symbol(date)
        stop_times %>%
                group_by(route_id) %>%
                select(route_id, date) %>%
                mutate(delay=dateColumn) %>%
                filter(delay != "NA") %>%
                summarise(num_departures=n(),
                          mean=mean(delay, na.rm=TRUE), 
                          sample_sd=sd(delay, na.rm=TRUE)) %>%
                arrange(desc(num_departures), mean) %>%
                as.data.frame()
})
```

Kruskal-Wallis test on the bus route delays:
```{r, echo=FALSE}
renderPrint({
        date <- as.character(as.Date(input$dateInput3))
        dateColumn <- as.symbol(date)
        data <- stop_times %>%
                group_by(route_id) %>%
                select(date) %>%
                mutate(delay=dateColumn) %>%
                filter(delay != "NA") %>%
                select(delay) %>%
                as.data.frame()
        data$route_id <- as.factor(data$route_id)
        with(data, kruskal.test(delay, route_id))
})
```

Pairwise Wilcox Mann-Whitney U Test on bus delays to find significantly different routes
```{r, echo=FALSE}
renderPrint({
        date <- as.character(as.Date(input$dateInput3))
        dateColumn <- as.symbol(date)
        data <- stop_times %>%
                group_by(route_id) %>%
                select(date) %>%
                mutate(delay=dateColumn) %>%
                filter(delay != "NA") %>%
                select(delay) %>%
                as.data.frame()
        data$route_id <- as.factor(data$route_id)
        test <- with(data, pairwise.wilcox.test(delay, route_id, 
                                        p.adjust="bonferroni"))
        significant_routes <- melt(test[[3]]) %>%
                filter(value < 0.05 & value != "NA") %>%
                pull(X1) %>%
                unique()
        levels(significant_routes)
})
```

Because of the varying sample sizes, most of the routes become "significant" even though they 
may not be.

### Delays by Date
We can figure out if some delays are unexpectedly high at certain times of day.
```{r, echo=FALSE}
getBusRoutes <- reactive({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        routes <- stop_times %>%
                        group_by(route_id) %>%
                        select(route_id, date) %>%
                        mutate(delay=dateColumn) %>%
                        filter(delay != "NA") %>%
                        ungroup() %>%
                        distinct(route_id) %>%
                        pull()
        return(routes)
})

inputPanel(
        dateInput("dateInput4",
                  label="Choose a Date:",
                  min="2018-12-01", max=Sys.Date(),
                  value=Sys.Date()),
        renderUI({
                selectInput(inputId="busRoute2", 
                            label="Choose a Bus Route:", 
                            choices=getBusRoutes(),
                            selected="ILLINI")
        })
)

renderPlotly({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        hours <- stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                pull(arrival_time) %>% 
                stringr::str_sub(1, 2) %>% 
                as.factor()
        delays <- stop_times %>%
                filter(route_id==input$busRoute2) %>%
                select(date) %>%
                mutate(delay=dateColumn / 60) %>%
                filter(delay != "NA") %>%
                select(delay)
        data.frame(delay=delays, hour=hours) %>%
                ggplot(aes(x=hour, y=delay)) + 
                        geom_violin() +
                        labs(x="Hour of Day", y="Delay (minutes)",
                             title=paste("Delays by hour of day for", input$busRoute2))
})
```

Is there a statistical difference between the distributions of these times? Since these don't 
seem to be normally distributed, we can use a nonparametric test to figure this out. (If this 
fails, then there are probably too little data points to compare)
```{r, echo=FALSE}
renderPrint({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        hours <- stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                pull(arrival_time) %>% 
                stringr::str_sub(1, 2) %>% 
                as.factor()
        delays <- stop_times %>%
                filter(route_id==input$busRoute2) %>%
                select(date) %>%
                mutate(delay=dateColumn / 60) %>%
                filter(delay != "NA") %>%
                pull(delay)
        data <- data.frame(delay=delays, hour=hours)
        with(data, kruskal.test(delay, hour))
})
```

Pairwise Wilcox Mann-Whitney U Test on bus delays to find significantly different times of day
```{r, echo=FALSE}
renderPrint({
        date <- as.character(as.Date(input$dateInput4))
        dateColumn <- as.symbol(date)
        hours <- stop_times %>%
                filter(route_id==input$busRoute2 & dateColumn != "NA") %>%
                pull(arrival_time) %>% 
                stringr::str_sub(1, 2) %>% 
                as.factor()
        delays <- stop_times %>%
                filter(route_id==input$busRoute2) %>%
                select(date) %>%
                mutate(delay=dateColumn / 60) %>%
                filter(delay != "NA") %>%
                pull(delay)
        data <- data.frame(delay=delays, hour=hours)
        test <- with(data, pairwise.wilcox.test(delay, hour, p.adjust.method = "bonferroni"))
        melt(test[[3]]) %>%
                filter(value < 0.05 & value != "NA") %>%
                arrange(X1) %>%
                pull(X1) %>%
                unique()
})
```

